---
title: "Data Cleaning"
author: "Geoff Willard"
date: "2/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = FALSE, message = FALSE}
library(dplyr)
library(tidyr)
```

package::function()
stats::filter()

# Read in data
We are using data from 
[Mike Byerly. Alaska commercial salmon catches by management region (1886- 1997). Gulf of Alaska Data Portal. df35b.304.2.](https://knb.ecoinformatics.org/view/df35b.304.2)

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"), stringsAsFactors = F)

head(catch_original)
```


# Clean and reshape
Remove marginal sum and notes columns using 'select()'

```{r}
catch_clean <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook))

head(catch_clean)
```

##QA: Quality Assurance
Find where as.numeric couldn't convert Chinook to a number

```{r}
i <- which(is.na(catch_clean$Chinook))
i
```

```{r}
catch_original[i, ]
```

Reshape from wide to tall format using 'pivot_longer'

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "catch") %>% 
  rename(catch_thousands = catch) %>% 
  mutate(catch = catch_thousands * 1000) %>% 
  select(-catch_thousands)

head(catch_long)
```

Example with pivot wider

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = Year, 
              values_from = catch)

head(catch_wide)
```


##group_by and summarize

```{r}
mean_region <- catch_long %>% 
  group_by(Region, species) %>% 
  summarise(catch_mean = mean(catch),
            num_obs = n())

head(mean_region)
```

```{r}
mean_species <- catch_long %>% 
  group_by(species) %>% 
  summarise(mean_catch = mean(catch),
            num_obs = n())

head(mean_species)
```

What year was the total salmon catch the highest?
```{r}
annual_catch <- catch_long %>% 
  group_by(Year) %>% 
  summarize(total_catch = sum(catch)) %>% 
  arrange(-total_catch)

annual_catch
```


What year had the highest SOCKEYE catch?
```{r}
annual_sockeye <- catch_long %>% 
  filter(species == "Sockeye") %>% 
  group_by(Year) %>% 
  summarize(total_catch = sum(catch)) %>% 
  arrange(-total_catch)

head(annual_sockeye)
```


# Join to regions table

Read in regions data
```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", stringsAsFactors = FALSE) %>% 
  select(code, mgmtArea)

head(region_defs)
```

```{r}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))

head(catch_joined)
```

